---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load the data 
```{r echo=TRUE}
originalData <- read.csv("activity.csv")
```

A portion of the original dataset
```{r echo=FALSE}
print(originalData[1:20,])
```


## What is mean total number of steps taken per day?

1. A dataset containing the total number of steps taken each day is created.

```{r echo=TRUE}
  dailyStepSum <- aggregate(originalData$steps, list(originalData$date), sum)
  ```
   A portion of the new dataset is as follows:
  ```{r echo=FALSE}
  colnames(dailyStepSum) <- c("Date", "Steps")
  print(dailyStepSum[1:20,])
```

2. A histogram of the above data is created as a form of visual representation.

```{r fig.width=15, echo=TRUE, tidy=FALSE}
  with(dailyStepSum, {
      par(oma=c(2,0,0,0), mar=c(6.75,6.75,3,0), mgp=c(5.75,0.75,0), las=2)
      barplot(
        height=Steps,
        main="Graph of Total Steps taken per Day",
        xlab="Dates",
        ylab="Steps per Day",
        names.arg=Date,
        space=c(0)
      )
  })
```
3. Calculate the mean and median values (ignoring NA values) using the above dataset.

  1. Mean
    ```{r echo=TRUE}
      dailyStepMean <- mean(dailyStepSum$Steps, na.rm=TRUE)
      ```
      ```{r echo=FALSE}
      print(dailyStepMean)
    ```
  2. Median
    ```{r echo=TRUE}
      dailyStepMedian <- median(dailyStepSum$Steps, na.rm=TRUE)
    ```
    ```{r echo=FALSE}
      print(dailyStepMedian)
    ```


## What is the average daily activity pattern?
1. Generate the mean (average) number of steps taken (ignoring NA values) for each 5-minute interval, itself averaged across all days.
  
```{r echo=TRUE, tidy=FALSE}
  intervalSteps <- aggregate(
      data=originalData,
      steps~interval,
      FUN=mean,
      na.action=na.omit
  )
  colnames(intervalSteps) <- c("Interval", "AvgStepsAvgAcrossDay")
```
   A portion of the new dataset is as follows:
```{r echo=FALSE}
  print(intervalSteps[1:20,])
```
  
2. A Time-Series plot is created from the above datase

```{r fig.width=15, echo=TRUE, tidy=FALSE}
  with(intervalSteps, {
      plot(
        x=Interval,
        y=AvgStepsAvgAcrossDay,
        type="l",
        main="Time-Series of Average Steps against Interval",
        xlab="5-minute Interval",
        ylab="Average Steps, Average across all Days"
        
      )
  })
```
3. Finding the 5-minute interval with the maximum number of steps

```{r echo=TRUE, tidy=FALSE}
  intervalMax <- intervalSteps[intervalSteps$AvgStepsAvgAcrossDay==max(intervalSteps$AvgStepsAvgAcrossDay),]
```
```{r echo=FALSE}
  print(intervalMax)
```

## Imputing missing values

1. Total number of rows with NA values in original data.

```{r echo=TRUE}
  countNA <- nrow(subset(originalData, is.na(originalData$steps)))
```
```{r echo=FALSE}
  print(countNA)
```

2. The average 5-minute interval values from the prevous section is used to replace the NA values of the original data and a new dataset will be generated from the latter.

 Decimal values will be rounded up to a whole number.
 
```{r echo=TRUE, tidy=FALSE}
  stepValues <- data.frame(originalData$steps)
  stepValues[is.na(stepValues),] <- ceiling(tapply(X=originalData$steps,INDEX=originalData$interval,FUN=mean,na.rm=TRUE))
  newData <- cbind(stepValues, originalData[,2:3])
  colnames(newData) <- c("Steps", "Date", "Interval")
```
  
  A portion of the new dataset is as follows:
```{r echo=FALSE}
  print(newData[1:20,])
```

3. The total number of steps taken each day is generated using this new dataset.

```{r echo=TRUE}
  newDailyStepSum <- aggregate(newData$Steps, list(newData$Date), sum)
```
   A portion of the new dataset is as follows:
```{r echo=FALSE}
  colnames(newDailyStepSum) <- c("Date", "Steps")
  print(newDailyStepSum[1:20,])
```

4. A histogram of the above data is created as a form of visual representation.

```{r fig.width=15, echo=TRUE, tidy=FALSE}
  with(newDailyStepSum, {
      par(oma=c(2,0,0,0), mar=c(6.75,6.75,3,0), mgp=c(5.75,0.75,0), las=2)
      barplot(
        height=Steps,
        main="Graph of Total Steps taken per Day",
        xlab="Dates",
        ylab="Steps per Day",
        names.arg=Date,
        space=c(0)
      )
  })
```

## Are there differences in activity patterns between weekdays and weekends?
1.  A new column indicating whether the date is a weekday or a weekend is added to the new dataset created in the previous section.

```{r echo=TRUE}
  dateDayType <- data.frame(sapply(X=newData$Date, FUN=function(day) {
    if (weekdays(as.Date(day)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) {
      day <- "weekday"
    }
    else {
      day <- "weekend"
    } 
  }))
  
  newDataWithDayType <- cbind(newData, dateDayType)
  
  colnames(newDataWithDayType) <- c("Steps", "Date", "Interval", "DayType")
```
  
   A portion of this dataset is as follows:
```{r echo=FALSE}
  print(newDataWithDayType[1:20,])
```
2. The data is then separated into weekday or weekend and the mean (average) number of steps taken for each 5-minute interval, itself averaged across all weekday days or weekend days is calculated.

```{r echo=TRUE, tidy=FALSE}
  dayTypeIntervalSteps <- aggregate(
      data=newDataWithDayType,
      Steps ~ DayType + Interval,
      FUN=mean
  )
```  
   A portion of the dataset is as follows:
```{r echo=FALSE}
  print(dayTypeIntervalSteps[1:20,])
```

3. Finally, a panel plot of both weekend and weekday graphs is generated.

```{r fig.width=15, echo=TRUE, tidy=FALSE}
  library("lattice")
  
  xyplot(
      type="l",
      data=dayTypeIntervalSteps,
      Steps ~ Interval | DayType,
      xlab="Interval",
      ylab="Number of steps",
      layout=c(1,2)
  )
```

